Structured Versus Decorated Cospans

### 2020-12-06 cospans.tex
polished up Kenny's argument that open dynamical systems aren't a structured cospan double category

Recently Pollard and the first author showed how to construct an `open' dynamical system from an `open' Petri net with rates, and define a functor from a category with open Petri nets with rates as morphisms to one with open dynamical systems as morphisms \cite{BP}.
 Later, structured cospans were used to promote the first of these categories to a double category \cite[Sec.\ 6.16]{BC}. Here we show that the second of these categories, with open dynamical systems as morphisms, \emph{cannot} be made into a double category using structured cospans---but we \emph{can} do it using decorated cospans!

@@

	\end{pgfonlayer}
	\begin{pgfonlayer}{edgelayer}
		\draw [style=inarrow, bend left =10, looseness=1.00] (S) to (tau1);
		\draw [style=inarrow, bend left=15, looseness=1.00] (I) to (tau1);
		\draw [style=inarrow, bend left=25, looseness=1.00] (tau1) to (I);
		\draw [style=inarrow, bend left=40, looseness=1.00] (tau1) to (I);
	       \draw [style=inarrow, bend right=25, looseness=1.00] (I) to (tau2);
             \draw [style=inarrow, bend right=15, looseness=1.00] (tau2) to (R); 
\path[color=purple, very thick, shorten >=2pt, shorten <=2pt, ->, >=stealth] (1a) edge (1'a);
\path[color=purple, very thick, shorten >=2pt, shorten <=2pt, ->, >=stealth] (1b) edge (1'b);
\path[color=purple, very thick, shorten >=2pt, shorten <=2pt, ->, >=stealth] (2) edge (2');	
\path[color=purple, very thick, shorten >=2pt, shorten <=2pt, ->, >=stealth] (3) edge (3');
		\draw [style=simple] (ATL.center) to (ATR.center);
		\draw [style=simple] (ATR.center) to (ABR.center);
		\draw [style=simple] (ABR.center) to (ABL.center);
		\draw [style=simple] (ABL.center) to (ATL.center);
		\draw [style=simple] (MTL.center) to (MTR.center);
		\draw [style=simple] (MTR.center) to (MBR.center);
		\draw [style=simple] (MBR.center) to (MBL.center);
		\draw [style=simple] (MBL.center) to (MTL.center);
	\end{pgfonlayer}
\end{tikzpicture}
\]
It is an open Petri net where the transitions are labelled with rate constants $r_1, r_2 > 0$.
Here is the corresponding open dynamical system:
\[   \begin{array}{ccl} \displaystyle{\frac{dS(t)}{dt}} &=& -r_1 \, S(t)I(t)  + I_1(t) + I_2(t) \\ \\
\displaystyle{\frac{dI(t)}{dt}}  &=& r_1\, S(t)I(t) - r_2 \, I(t) + I_3(t)  \\  \\
\displaystyle{\frac{dR(t)}{dt}}  &=&   r_2 \, I(t)  - O_1(t).
\end{array}
\]
If we drop the `inflow' and `outflow' terms $I_i$ and $O_i$ we obtain a system of coupled nonlinear first-order ordinary differential equations.   In fact these equations are a famous model of infectious disease, the `SIR model', where $S(t)$, $I(t)$ and $R(t)$ describe the populations of susceptible, infected and recovered individuals, respectively.   The inflow and outflow terms allow individuals to enter or leave the population.   This in turn lets us couple the SIR model to other models, and build complex models from smaller pieces.  
Recently Halter, and Patterson \cite{HP} implemented this idea in software using structured cospans.  They illustrated the use of this tool by rebuilding part of the model that the UK has been using to make policy decisions about COVID-19.

Now let us turn to the mathematics.  A \define{Petri net with rates} is a Petri net $s,t \maps T \to \N[S]$ together with a function $r \maps T \to (0,\infty)$ assigning to each transition $\tau \in T$ a positive real number called its \define{rate constant}.  There is a category $\Petri_r$ whose objects are Petri nets with rates, where a morphism from 
\[   \xymatrix{ (0,\infty) & T \ar[l]_-r \ar@<-.5ex>[r]_-t \ar@<.5ex>[r]^-s & \N[S] }\]
 to 
 \[   \xymatrix{ (0,\infty) & T' \ar[l]_-r \ar@<-.5ex>[r]_-t \ar@<.5ex>[r]^-s & \N[S'] }\]
is a morphism of the underlying Petri nets such that the following diagram also commutes:
\[
\begin{tikzpicture}[scale=1.5]
\node (E) at (3,-0.5) {$(0,\infty)$};
\node (G) at (4,0) {$T$};
\node (G') at (4,-1) {$T'$};
\path[->,font=\scriptsize,>=angle 90]
(G) edge node[right]{$f$} (G')
(G) edge node[above]{$r$} (E)
(G') edge node[below]{$r'$} (E);
\end{tikzpicture}
\]
There is a functor $R \maps \Petri_r \to \Set$ sending any Petri net with rates to its set of places, and this has a left adjoint $L \maps \Set \to \Petri_r$ sending any set $S$ to the Petri net with $S$ as its set of species and no transitions \cite[Lemma 6.18]{BC}.   Since $\Petri_r$ has finite colimits \cite[Lemma 6.19]{BC}, it follows that there is a symmetric monoidal double category ${}_L \lCsp(\Petri_r)$ where:
\begin{itemize}
\item objects are finite sets,
\item vertical 1-morphisms are functions,
\item horizontal 1-cells are \define{open Petri nets with rates}:
\[
\begin{tikzpicture}[scale=1.5]
\node (D) at (-3,0) {$L(X)$};
\node (E) at (-2,0) {$P$};
\node (F) at (-1,0) {$L(Y)$};
\path[->,font=\scriptsize,>=angle 90]
(D) edge node [above] {$i$} (E)
(F) edge node [above] {$o$} (E);
\end{tikzpicture}
\]
\item 2-morphisms are diagrams in $\Petri_r$ of the form:
\[
\begin{tikzpicture}[scale=1.5]
\node (E) at (3,0) {$L(X_1)$};
\node (F) at (5,0) {$L(Y_1)$};
\node (G) at (4,0) {$P_1$};
\node (E') at (3,-1) {$L(X_2)$};
\node (F') at (5,-1) {$L(Y_2)$.};
\node (G') at (4,-1) {$P_2$};
\path[->,font=\scriptsize,>=angle 90]
(F) edge node[above]{$O_1$} (G)
(E) edge node[left]{$L(f)$} (E')
(F) edge node[right]{$L(g)$} (F')
(G) edge node[left]{$\alpha$} (G')
(E) edge node[above]{$I_1$} (G)
(E') edge node[above]{$I_2$} (G')
(F') edge node[above]{$O_2$} (G');
\end{tikzpicture}
\]
\end{itemize}

@@

Using Fong's original approach to decorated cospans, Pollard and the first author constructed a symmetric monoidal category for which the morphisms are open dynamical systems \cite[Theorem 17]{BP}.  This category is constructed from a symmetric lax monoidal functor 
$D \maps \Fin\Set \to \Set$ such that:
\begin{itemize}
\item $D$ maps any finite set $S$ to 
\[ D(S) = \{ v \maps \R^S \to \R^S : \; v \textrm{ is algebraic}  \}. \] 
\item $D$ maps any function $f \maps S \to S'$ between finite sets to the function $D(f) \maps D(S) \to D(S')$ given as follows:
\[ D(f)(v) = f_* \circ v \circ f^* \]
where the pullback $ f^* \maps \R^{S'} \to \R^S $ is given by
\[ f^*(c)(\sigma) = c(f(\sigma)) \] 
while the pushforward $ f_* \maps \R^{S} \to \R^{S'} $ is given by
\[ f_*(c)(\sigma') = \sum_{ \{ \sigma \in S : f(\sigma) = \sigma' \} } c(\sigma). \]
\end{itemize}
The functorality of $D$ is proved in \cite[Lemma 15]{BP} while the lax symmetric monoidal stucture is given in Lemma 16.

@@

We now show that while open dynamical systems are morphisms in a decorated cospan category, they do not arise from a structured cospan category via \cref{thm:equiv}.  To do this, recall that in this result we start with the data required for a decorated cospan category, namely a symmetric lax monoidal pseudofunctor $F \maps (\A,+) \to (\Cat,\times)$, and show the resulting opfibration $U \maps \inta F \to \A$ has a left adjoint $L \maps \A \to \X$.   
We shall show that in the case at hand, where $F = D$, the opfibration $U \maps \inta F \to \A$ does \emph{not} have a left adjoint.

Let $\mathsf{Dynam}$ be the category for which:
\begin{itemize}
\item{an object is a pair $(s,v_s)$ where $s$ is a finite set and $v_s$ is an algebraic vector field $v_s \colon \lR^s \to \lR^s$ on the finite set $s$, and}
\item{a morphism from $(s,v_s)$ to $(s',v_{s'})$ is a function $f \colon s \to s'$ such that $v_{s'} = f_* v_s f^*$.}
\end{itemize}

A well-known result is the following:
\begin{lem}
A functor $U \colon \A \to \X$ admits a left adjoint if and only if for every $x \in \X$, the comma category $x \downarrow U$ has an initial object.
\end{lem}
Using this result, one can show that the forgetful functor $U \colon \mathsf{Dynam} \to \Fin\Set$ is not a right adjoint. In particular, by taking $x$ to be the empty set, the comma category $\emptyset \downarrow U$ has:
\begin{itemize}
\item{objects given as pairs $((s,v_s),! \colon \emptyset \to s)$, and}
\item{a morphism from $((s,v_s),! \colon \emptyset \to s)$ to $((s',v_{s'}),! \colon \emptyset \to s')$ is a morphism $f \colon (s,v_s) \to (s',v_{s'})$ in $\mathsf{Dynam}$ such that $U(f) \circ ! = !$ in $\Fin\Set$, a condition which amounts to simply being a morphism in $\mathsf{Dynam}$.}
\end{itemize}
In particular, the object $((\emptyset,v_\emptyset),! \colon \emptyset \to \emptyset)$ resides in the comma category $\emptyset \downarrow U$, where $v_\emptyset$ is the only possible vector field on the empty set --- namely, the dengerate vector field. Given any other object $((s,v_s),! \colon \emptyset \to s)$ in $\emptyset \downarrow U$, there exists no morphisms $! \colon \emptyset \to s$ from the first object to the second unless $s$ is also the empty set as the vector field $$!_* v_\emptyset !^*$$will also always be degenerate and not equal to $v_s$. Clearly there are also no morphisms in the other direction for $s \neq \emptyset$ as $f$ must be a function. Thus the forgetful functor $U$ has no left adjoint $L \colon \Fin\Set \to \mathsf{Dynam}$, and hence we cannot construct a structured cospan category ${_L \Csp(\mathsf{Dynam})}$ whose morphisms are equivalence classes of open dynamical systems of this type.

\section{Conclusions}\label{sec:conclusions}
%%% Christina: Or future directions or something? Unless we at some point conclude about something for the two frameworks, i.e they are close enough that we can choose which one to work with according to our application or something? I did something along those lines at the very end....

>>>>

Pollard and the first author have used decorated cospans to construct an `open' dynamical system from any `open' Petri net with rates, thus defining a functor from a category with open Petri nets with rates as morphisms to one with open dynamical systems as morphisms \cite{BP}. Later, structured cospans were used to promote the first of these categories to a double category \cite[Sec.\ 6.16]{BC}. Here we show that the second of these categories, with open dynamical systems as morphisms, \emph{cannot} be made into a double category using structured cospans.  However, we \emph{can} do it using decorated cospans.

@@

	\end{pgfonlayer}
	\begin{pgfonlayer}{edgelayer}
		\draw [style=inarrow, bend left =10, looseness=1.00] (S) to (tau1);
		\draw [style=inarrow, bend left=15, looseness=1.00] (I) to (tau1);
		\draw [style=inarrow, bend left=25, looseness=1.00] (tau1) to (I);
		\draw [style=inarrow, bend left=40, looseness=1.00] (tau1) to (I);
	       \draw [style=inarrow, bend right=25, looseness=1.00] (I) to (tau2);
             \draw [style=inarrow, bend right=15, looseness=1.00] (tau2) to (R); 
\path[color=purple, very thick, shorten >=2pt, shorten <=2pt, ->, >=stealth] (1a) edge (1'a);
\path[color=purple, very thick, shorten >=2pt, shorten <=2pt, ->, >=stealth] (1b) edge (1'b);
\path[color=purple, very thick, shorten >=2pt, shorten <=2pt, ->, >=stealth] (2) edge (2');	
\path[color=purple, very thick, shorten >=2pt, shorten <=2pt, ->, >=stealth] (3) edge (3');
		\draw [style=simple] (ATL.center) to (ATR.center);
		\draw [style=simple] (ATR.center) to (ABR.center);
		\draw [style=simple] (ABR.center) to (ABL.center);
		\draw [style=simple] (ABL.center) to (ATL.center);
		\draw [style=simple] (MTL.center) to (MTR.center);
		\draw [style=simple] (MTR.center) to (MBR.center);
		\draw [style=simple] (MBR.center) to (MBL.center);
		\draw [style=simple] (MBL.center) to (MTL.center);
	\end{pgfonlayer}
\end{tikzpicture}
\]
It is an open Petri net where the transitions are labelled with rate constants $r_1, r_2 > 0$.
Here is the corresponding open dynamical system:
\[   \begin{array}{ccl} \displaystyle{\frac{dS(t)}{dt}} &=& -r_1 \, S(t)I(t)  + I_1(t) + I_2(t) \\ \\
\displaystyle{\frac{dI(t)}{dt}}  &=& r_1\, S(t)I(t) - r_2 \, I(t) + I_3(t)  \\  \\
\displaystyle{\frac{dR(t)}{dt}}  &=&   r_2 \, I(t)  - O_1(t).
\end{array}
\]
If we drop the `inflow' and `outflow' terms $I_i$ and $O_i$ we obtain a system of coupled nonlinear first-order ordinary differential equations.   In fact these equations are a famous model of infectious disease, the `SIR model', where $S(t)$, $I(t)$ and $R(t)$ describe the populations of susceptible, infected and recovered individuals, respectively.   The inflow and outflow terms allow individuals to enter or leave the population.   This in turn lets us couple the SIR model to other models, and build complex models from smaller pieces.  
Recently Halter and Patterson \cite{HP} implemented this idea in software using structured cospans.  They illustrated the use of this tool by rebuilding part of the COVID-19 model that the UK has been using to make policy decisions.

Now let us turn to the details.  A \define{Petri net with rates} is a Petri net $s,t \maps T \to \N[S]$ together with a function $r \maps T \to (0,\infty)$ assigning to each transition $\tau \in T$ a positive real number called its \define{rate constant}.  There is a category $\Petri_r$ whose objects are Petri nets with rates, where a morphism from 
\[   \xymatrix{ (0,\infty) & T \ar[l]_-r \ar@<-.5ex>[r]_-t \ar@<.5ex>[r]^-s & \N[S] }\]
 to 
 \[   \xymatrix{ (0,\infty) & T' \ar[l]_-r \ar@<-.5ex>[r]_-t \ar@<.5ex>[r]^-s & \N[S'] }\]
is a morphism of the underlying Petri nets such that the following diagram also commutes:
\[
\begin{tikzpicture}[scale=1.5]
\node (E) at (3,-0.5) {$(0,\infty)$};
\node (G) at (4,0) {$T$};
\node (G') at (4,-1) {$T'$};
\path[->,font=\scriptsize,>=angle 90]
(G) edge node[right]{$f$} (G')
(G) edge node[above]{$r$} (E)
(G') edge node[below]{$r'$} (E);
\end{tikzpicture}
\]
There is a functor $R \maps \Petri_r \to \Set$ sending any Petri net with rates to its set of places, and this has a left adjoint $L \maps \Set \to \Petri_r$ sending any set $S$ to the Petri net with $S$ as its set of species and no transitions \cite[Lemma 6.18]{BC}.   Since $\Petri_r$ has finite colimits \cite[Lemma 6.19]{BC}, it follows that there is a symmetric monoidal double category ${}_L \lCsp(\Petri_r)$ where:
\begin{itemize}
\item objects are finite sets,
\item vertical 1-morphisms are functions,
\item horizontal 1-cells are \define{open Petri nets with rates}:
\[
\begin{tikzpicture}[scale=1.5]
\node (D) at (-3,0) {$L(X)$};
\node (E) at (-2,0) {$P$};
\node (F) at (-1,0) {$L(Y)$};
\path[->,font=\scriptsize,>=angle 90]
(D) edge node [above] {$i$} (E)
(F) edge node [above] {$o$} (E);
\end{tikzpicture}
\]
\item 2-morphisms are diagrams in $\Petri_r$ of the form:
\[
\begin{tikzpicture}[scale=1.5]
\node (E) at (3,0) {$L(X_1)$};
\node (F) at (5,0) {$L(Y_1)$};
\node (G) at (4,0) {$P_1$};
\node (E') at (3,-1) {$L(X_2)$};
\node (F') at (5,-1) {$L(Y_2)$.};
\node (G') at (4,-1) {$P_2$};
\path[->,font=\scriptsize,>=angle 90]
(F) edge node[above]{$O_1$} (G)
(E) edge node[left]{$L(f)$} (E')
(F) edge node[right]{$L(g)$} (F')
(G) edge node[left]{$\alpha$} (G')
(E) edge node[above]{$I_1$} (G)
(E') edge node[above]{$I_2$} (G')
(F') edge node[above]{$O_2$} (G');
\end{tikzpicture}
\]
\end{itemize}

@@

Using Fong's original approach to decorated cospans, Pollard and the first author constructed a symmetric monoidal category for which the morphisms are open dynamical systems \cite[Theorem 17]{BP}.  This category is constructed from a symmetric lax monoidal functor 
$D \maps \Fin\Set \to \Set$ such that:
\begin{itemize}
\item $D$ maps any finite set $S$ to 
\[ D(S) = \{ v \maps \R^S \to \R^S : \; v \textrm{ is algebraic}  \}. \] 
\item $D$ maps any function $f \maps S \to S'$ between finite sets to the function $D(f) \maps D(S) \to D(S')$ given as follows:
\[ D(f)(v) = f_* \circ v \circ f^* \]
where the pullback $ f^* \maps \R^{S'} \to \R^S $ is given by
\[ f^*(c)(\sigma) = c(f(\sigma)) \] 
while the pushforward $ f_* \maps \R^{S} \to \R^{S'} $ is given by
\[ f_*(c)(\sigma') = \sum_{ \{ \sigma \in S : f(\sigma) = \sigma' \} } c(\sigma). \]
\end{itemize}
The functorality of $D$ is proved in \cite[Lemma 15]{BP} while the lax symmetric monoidal stucture is given in Lemma 16 of that paper.

@@

We now show that while open dynamical systems are morphisms in a decorated cospan category, this does not arise from a structured cospan category via \cref{thm:equiv}.  To do this, recall that in that theorem we start with the data required to build a decorated cospan category, namely a symmetric lax monoidal pseudofunctor $F \maps (\A,+) \to (\Cat,\times)$, and show that under certain conditions the resulting opfibration $U \maps \inta F \to \A$ has a left adjoint $L \maps \A \to \X$.   We then obtain an equivalence between decorated and structured cospan double categories, $F \lCsp \simeq {}_L \Csp(\X)$.   We now show that in the case at hand, where $F = D$, $U$ does \emph{not} have a left adjoint.  Thus, the conditions of \cref{thm:equiv} do not hold in this case.

Taking $D$ as above, it is easy to see that in the category $\inta D$
\begin{itemize}
\item an object is a pair $(S,v)$ where $S$ is a finite set and $v$ is an algebraic vector field $v \maps \R^S \to \R^S$,
\item a morphism from $(S,v)$ to $(S',v')$ is a function $f \maps S \to S'$ such that $v' = f_* \circ v \circ f^*$.
\end{itemize}
with the usual composition of functions.   The forgetful functor $U \maps \inta D \to \Fin\Set$ acts as follows:
\begin{itemize}
\item on objects, $D(S,v) = S$,
\item on morphisms, $D(f) = f$.
\end{itemize}

To show that $U$ does not have a left adjoint, we use the following well-known result:
\begin{lem} \label{lem:initial}
A functor $U \maps \A \to \X$ admits a left adjoint if and only if for every $x \in \X$, the comma category $x \downarrow U$ has an initial object.
\end{lem}
Because the empty set is initial in $\Fin\Set$, the comma category 
$\emptyset \downarrow U$ is just $\inta D$.  This contains an object $(\emptyset, v_\emptyset)$, where $v_\emptyset$ is the only possible vector field on $\R^\emptyset$, namely, the zero vector field.   The only object in $\inta D$ with any morphisms to $(\emptyset, v_\emptyset)$ is $(\emptyset, v_\emptyset)$ itself, so no other object can be initial.  However $(\emptyset, v_\emptyset)$ is not initial either, because it has no morphisms to an object $(S,v)$ unless $v$ is the zero vector field on $\R^S$.  Thus by \cref{lem:initial}, $U$ does not have a left adjoint.

\section{Conclusions}\label{sec:conclusions}
